Tests of measurement invariance via stochastic processes: {S}ome properties

### 2013-08-05 frontiers/frontiers.Rnw
turn off the signif codes in anova. use sctest.lavaan function to get the continuous and ordinal statistics
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@229 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

## auxiliary code
source("simall.R")
source("mz.R")

@@

<<theta estimate>>=
# estimate theta 
library(lavaan)
theta<-cfa(
  'f1=~gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5',
  data=yg,group="agegroup",meanstructure=TRUE,
  group.equal="loadings")
@
Then we can get estimate for group A and group B. 
<<theta A and theta B estimate>>=
# estimate theta A,B 
thetaAB<-cfa(
  'f1=~gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5',
  data=yg,group="agegroup",meanstructure=TRUE)
@
In the end, we can get the LRT results by calling the measurement invariance function in lavaan. 
<<LRT>>=
# LRT
LRT<-anova(theta, thetaAB)
LRT
@
\subsection{Extensions for Unknown Group}
The multiple group test performs typically well when the $V$ is a relatively-small number of categories and the sample size is not too large (LRT is sensitive to large sample and will return "false" significance results, details see next section). However, when $V$ (in this case, age) continuous variable, however, multiple-group analysis usually cannot be performed because there are no existing groups. Instead, we can follow the procedure proposed by Ed Merkle and Zeileis. Specifically, we can fit a model whose parameters are restricted to be equal across across all individuals and then examine how individuals' scores $s(\hat{theta};x_{i})$ fluctuate with their values of $V$. To formalize the idea, we assume that the observations are ordered w.r.t. $V$ and define the $k$ dimensional cumulative score process as 
\begin{equation}
\label{equ:B}
B(t;\hat{\theta})=\hat{I}^{-1/2}n^{-1/2}\overset{\lfloor nt \rfloor}{\underset{i=1}{\Sigma}}s(\hat{\theta};x_{i}) (0 \leq t \leq 1)
\end{equation}
where $\lfloor nt \rfloor$ is the integer part of nt and $\hat{I}$ is some consistent estimate of the covariance matrix of the scores,e.g. information matrix. ~\ref{equ:B} simultaneously accounts for the ordering of individuals w.r.t. $V$ and decorrelates the scores between $k$ components of the parameters in the model. 
Further, there exists a functional central limit theorem that allows us to make formal inference with this cumulative score process. Assuming that individuals are independent and the usual ML regularity condition hold, it is possible to show that(addref) 
\begin{equation}
B(\cdot;\hat{\theta})\overset{d}{\rightarrow} B^{0}(\cdot)
\end{equation} 
where $\overset{d}{\rightarrow}$ denotes convergence in distribution and $B^{0}(\cdot)$ is a $k$ dimensional Brownian bridge. Thus, we can construct tests of measurement invariance by comparing the behavior of cumulative score process to that of a Brownian bridge. In practice, there are three test statistics can be used: double maximum, CvM and maxLM. 
\begin{equation}
DM={\underset{i=1,...,n}{\max}}{\underset{j=1,...,k}{\max}}|B(\hat{\theta})_{ij}|
\end{equation}
\begin{equation}
CvM=n^{-1}{\underset{i=1,...,n}{\Sigma}}{\underset{j=1,...,k}{\Sigma}}B(\hat{\theta_{ij}})^{2}
\end{equation}
\begin{equation}
maxLM={\underset{i=1,...,n}{\max}}{\frac{i}{n}(1-\frac{i}{n})}^{-1}{\underset{j=1,...,k}{\Sigma}}B(\hat{\theta}_{ij})^{2})
\end{equation}

Specifically, we will use the previous example, except changing categorical agegroup to continuous age.
Like before, after read in data, we fit the model in lavaan package. 
<<theta estimate>>=
library(lavaan)
theta<-cfa(
  'f1=~gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5',
  data=yg,group="agegroup",meanstructure=TRUE,
  group.equal="loadings")
@
Then we will use the gefp function in strucchange package to get the $B(\hat{\theta}_{ij})$. Then we can get the three test statistics' significance by using sctest function. 
<<Get >>=
gefp1<-gefp(theta,fit=NULL,order.by=as.numeric(yg$agegroup),
       vcov=info_full,sandwich=FALSE,parm=1:4)
pval=rep(NA,3)
pval[1]<-sctest(gefp1,functional=maxBB)$p.value
pval[2]<-sctest(gefp1,functional=meanL2BB)$p.value
pval[3]<-sctest(gefp1,functional=supLM(0.1))$p.value
pval
@

\subsection{Ordinal Test}
However, in this dataset, $V$ is actually ordinal instead of continuous. There is only a partial ordering of individuals with respect to $V$,i.e. observations with the same level of $V$ have no unique ordering. The ordinal statistics proposed here are similar to those described in Equation and above.

\SweaveOpts{engine = R, eps = FALSE, echo = FALSE, pdf=TRUE}
\section{Simulation 1}

>>>>

## auxiliary code
source("simall.R")
source("mz.R")
source("../www/efpFunctional-cat.R")
source("../pkg/sctest.lavaan.R")

@@

<<theta estimate>>=
# estimate theta 
library(lavaan)
theta<-cfa(
  'f1=~gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5',
  data=yg,group="agegroup",meanstructure=TRUE,
  group.equal="loadings")
@
Then we can get estimate for group A and group B. 
<<theta A and theta B estimate>>=
# estimate theta A,B 
thetaAB<-cfa(
  'f1=~gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5',
  data=yg,group="agegroup",meanstructure=TRUE)
@
In the end, we can get the LRT results by calling the measurement invariance function in lavaan. 
<<LRT>>=
# LRT
LRT<-as.matrix(anova(theta, thetaAB))
print(LRT,na.print="", quote=FALSE)
@
\subsection{Extensions for Unknown Group}
The multiple group test performs typically well when the $V$ is a relatively-small number of categories and the sample size is not too large (LRT is sensitive to large sample and will return "false" significance results, details see next section). However, when $V$ (in this case, age) continuous variable, however, multiple-group analysis usually cannot be performed because there are no existing groups. Instead, we can follow the procedure proposed by Ed Merkle and Zeileis. Specifically, we can fit a model whose parameters are restricted to be equal across across all individuals and then examine how individuals' scores $s(\hat{theta};x_{i})$ fluctuate with their values of $V$. To formalize the idea, we assume that the observations are ordered w.r.t. $V$ and define the $k$ dimensional cumulative score process as 
\begin{equation}
\label{equ:B}
B(t;\hat{\theta})=\hat{I}^{-1/2}n^{-1/2}\overset{\lfloor nt \rfloor}{\underset{i=1}{\Sigma}}s(\hat{\theta};x_{i}) (0 \leq t \leq 1)
\end{equation}
where $\lfloor nt \rfloor$ is the integer part of nt and $\hat{I}$ is some consistent estimate of the covariance matrix of the scores,e.g. information matrix. ~\ref{equ:B} simultaneously accounts for the ordering of individuals w.r.t. $V$ and decorrelates the scores between $k$ components of the parameters in the model. 
Further, there exists a functional central limit theorem that allows us to make formal inference with this cumulative score process. Assuming that individuals are independent and the usual ML regularity condition hold, it is possible to show that(addref) 
\begin{equation}
B(\cdot;\hat{\theta})\overset{d}{\rightarrow} B^{0}(\cdot)
\end{equation} 
where $\overset{d}{\rightarrow}$ denotes convergence in distribution and $B^{0}(\cdot)$ is a $k$ dimensional Brownian bridge. Thus, we can construct tests of measurement invariance by comparing the behavior of cumulative score process to that of a Brownian bridge. In practice, there are three test statistics can be used: double maximum, CvM and maxLM. 
\begin{equation}
DM={\underset{i=1,...,n}{\max}}{\underset{j=1,...,k}{\max}}|B(\hat{\theta})_{ij}|
\end{equation}
\begin{equation}
CvM=n^{-1}{\underset{i=1,...,n}{\Sigma}}{\underset{j=1,...,k}{\Sigma}}B(\hat{\theta_{ij}})^{2}
\end{equation}
\begin{equation}
maxLM={\underset{i=1,...,n}{\max}}{\frac{i}{n}(1-\frac{i}{n})}^{-1}{\underset{j=1,...,k}{\Sigma}}B(\hat{\theta}_{ij})^{2})
\end{equation}

Specifically, we will use the previous example, except changing categorical agegroup to continuous age.
Like before, after read in data, we fit the model in lavaan package. 
<<theta estimate>>=
library(lavaan)
theta<-cfa(
  'f1=~gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5',
  data=yg,group="agegroup",meanstructure=TRUE,
  group.equal="loadings")
@
Then we will use the gefp function in strucchange package to get the $B(\hat{\theta}_{ij})$. Then we can get the three test statistics' significance by using sctest function. 
<<Get Estimate>>=
gefp1<-gefp(theta,fit=NULL,order.by=as.numeric(yg$agegroup),
       vcov=info_full,sandwich=FALSE,parm=1:4)
pval=rep(NA,3)
pval[1]<-sctest(gefp1,functional=maxBB)$p.value
pval[2]<-sctest(gefp1,functional=meanL2BB)$p.value
pval[3]<-sctest(gefp1,functional=supLM(0.1))$p.value
pval
@
In applying the proposed continuous test, the $p$ value for CvM and maxBB $p$ value is larger than likelihood test. Besides, CvM $p$ value is not significant.

\SweaveOpts{engine = R, eps = FALSE, echo = TRUE, pdf=TRUE}
Alternatively, we could use sctest.lavaan function to get the results. 
<<Get Estimate Alternative Way>>=
DM<-sctest(theta, as.numeric(yg$agegroup),parm=1:4, stat="DM", fplot=TRUE) 
CvM<-sctest(theta, as.numeric(yg$agegroup),parm=1:4, stat="CvM", fplot=TRUE) 
#maxLM<-sctest(theta, as.numeric(yg$agegroup),parm=1:4, #stat="maxLM", fplot=TRUE) 
DM
CvM
#maxLM
@
\subsection{Ordinal Test}
However, in this dataset, $V$ is actually ordinal instead of continuous. There is only a partial ordering of individuals with respect to $V$,i.e. observations with the same level of $V$ have no unique ordering. The ordinal statistics proposed here are similar to those described in Equation and above, except that we focus on "bins" of individuals at each level of the ordinal variable. That is, instead of aggregating over all $i=1,...n$ individuals, we first compute cumulative proportions $t_{l}(l=1,...,m-1)$ associated with the first $m-1$ levels of $V$. We then aggregate the cumulative scores only over $i_{l}=\lfloor n \cdot t_{l} \rfloor$. Test statistics related can be written as 
\begin{eqnarray}
    \label{eq:wdm}
      \mathit{WDM}_o & = & \max_{i \in \{i_1, \dots, i_{m-1} \}} ~ \left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) \right\}^{-1/2}\max_{j = 1, \dots, k} | {\bm B}(\hat {\bm \theta})_{ij} |,\\
    \label{eq:maxlmo}
      \max \mathit{LM}_o & = & \max_{i \in \{i_1, \dots, i_{m-1} \}} ~\left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) \right\}^{-1}\sum_{j = 1, \dots, k} {\bm B}(\hat {\bm \theta})_{ij}^2,
\end{eqnarray}

The difference between the continuous test and ordinal test is that we don't force arbitary order on the observations within the same age groups. Therefore, this ordinal test should reflect more of the data. In terms of code, we still need the gefp function to get the B estimate. After that, we can use different functions to aggregate and scale B estimation. The function are called ordwmax and ol2bb1.fun, respectively. 
<<Ordinal Test>>=
gefp1<-gefp(theta,fit=NULL,order.by=as.numeric(yg$agegroup),
       vcov=info_full,sandwich=FALSE,parm=1:4)
pval.ord=rep(NA,2)
pval.ord[1]<-sctest(gefp1,functional=ordwmax(gefp1))$p.value
pval.ord[2]<-sctest(gefp1,functional=ordL2BB(gefp1))$p.value
pval.ord
@

<<Get Estimate Alternative Way for Ordinal>>=
WDMo<-sctest(theta, as.numeric(yg$agegroup),parm=1:4, stat="WDMo", fplot=TRUE) 
#maxLMo<-sctest(theta, as.numeric(yg$agegroup),parm=1:4, #stat="maxLMo", fplot=TRUE)
WDMo
#maxLMo
@

In employing the ordinal tests, we obtain $p=0.060$ and $p=0.096$. Both $p-$values are clearly larger than that of the likelihood ratio test and neigher is significant at $\alpha=0.5$

\section{Simulation 1}
